Effects of epidemiological structure on the transient evolution of HIV virulence

### 2016-10-24 ms/plos_latex.tex
a little more

Finally, the sensitivity plots (\figurename~\ref{fig:plot_sens}) show the effects 
of each parameter on the summary statistics. In almost every case the
effects of the parameters are monotonic; note that
the plot shows the effects of the \emph{unscaled} parameters, i.e.
before they have been calibrated to achieve a standard initial epidemic
growth rate.
Increases in the transmission rates ($\beta_P$, $\beta_D$)
and durations ($D_P$, $D_D$) in the primary and disease stages generally
decrease the equilibrium virulence, peak virulence, and peak time,
although the random and pair-formation+epc 
models have high, relatively
constant values with respect to these parameters
(because the

The partnership dissolution rate ($c$), which essentially
acts as a contact rate in the model,
increases virulence and peak time in almost all
cases, although the pair-formation+epc
model is again relatively insensitive.
The ratio of extra-pair to within-pair contact ($c_e/c_w$) affects
virulence in the instantaneous-switching+epc model, but not the pair-formation+epc
model (probably because the uncoupled individuals present in the pair-formation+epc
model make extra-pair contact by coupled individuals less important).
Surprisingly, neither of the pair-formation models is particularly sensitive to the 
rate of partnership formation ($\rho$); the rate of uncoupled contact
increases virulence and peak time in the pair-formation+epc model, 
which is the only model to which it applies.

@@

In the particular example of HIV virulence eco-evolutionary dynamics
and epidemiological, we reach the slightly ironic conclusion that the
effort put into building a more realistic model essentially cancels
out, leaving us in the same position as if we had ignored the problem
of epidemiological complexity entirely and used a naive random-mixing
contact model. In Herbeck \etal's \cite{herbeck_hiv_2014} network model of partnerships, the partnership duration is set to 1 day --- very unrealistic in epidemic terms, but perhaps
actually more true to real-world HIV epidemiological dynamics than a
model with realistic partnership durations that neglects extra-pair
contact \cite{herbeck2016evolution}. Making the model even more realistic
might make the random-mixing model less appropriate. For
example, our model forms partnerships randomly, and assumes that
extra-pair contact is randomly mixing across the population;
one could instead model extra-pair contact as arising from
multiple concurrent partnerships (some, such as contact with sex
workers, of very short duration) and/or more structured partnership
formation (by age, ethnicity, or behaviour group). The effects of
other realistic complications such as explicit modeling of two
sexes (both in contact structure and differential transmission
probabilities), temporal and spatial variation in epidemic processes,
or presence of genetic variation in hosts are harder to predict.

>>>>

Finally, the sensitivity plot (\figurename~\ref{fig:plot_sens}) shows the effects 
of each parameter on the summary statistics. In almost every case the
effects of the parameters are monotonic; note that
the plot shows the effects of the \emph{unscaled} parameters, i.e.
before they have been calibrated to achieve a standard initial epidemic
growth rate.
Increases in the transmission rates ($\beta_P$, $\beta_D$)
and durations ($D_P$, $D_D$) in the primary and disease stages generally
decrease the equilibrium virulence, peak virulence, and peak time,
although the random and pair-formation+epc 
models have high, relatively
constant values with respect to these parameters
(because the patterns are so similar across this set of parameters,
\figurename~\ref{fig:plot_sens} shows only $D_P$).

The partnership dissolution rate ($c$), which essentially
acts as a contact rate in the model,
increases virulence and peak time in almost all
cases, although the pair-formation+epc
model is again relatively insensitive.
The ratio of extra-pair to within-pair contact ($c_e/c_w$) affects
virulence in the instantaneous-switching+epc model, but not the pair-formation+epc
model (probably because the uncoupled individuals present in the pair-formation+epc
model make extra-pair contact by coupled individuals less important).
Surprisingly, once calibration
is taken into account, the remaining parameters have little effect overall.
The rate of partnership formation ($\rho$) 
has little impact on the models with finite pair-formation times.
The relative rate of uncoupled contact ($c_u/c_w$) slightly decreases the
minimum and equilibrium progression time and delays the peak time in the
pair-formation+epc model, but neither the uncoupled contact rate nor
the mean ($\mu$) or CV$^2$ of the number of non-cohabiting sexual
partners has much systematic effect in the heterogeneous model.

@@

In the particular example of HIV virulence eco-evolutionary dynamics
and the complexity of epidemiological structures
we reach the slightly ironic conclusion that the
effort put into building a more realistic model essentially cancels
out, putting us back where we started when used a naive random-mixing
contact model. In Herbeck \etal's \cite{herbeck_hiv_2014} network model of partnerships, the partnership duration is set to 1 day --- very unrealistic in epidemic terms, but perhaps
actually more true to real-world HIV epidemiological dynamics than a
model with realistic partnership durations that neglects extra-pair
contact \cite{herbeck2016evolution}. Making the model even more realistic
might make the random-mixing model less appropriate. For
example, our model forms partnerships randomly, and assumes that
extra-pair contact is randomly mixing across the population;
one could instead model extra-pair contact as arising from
multiple concurrent partnerships (some, such as contact with sex
workers, of very short duration) and/or more structured partnership
formation (by age, ethnicity, or behaviour group). The effects of
other realistic complications such as explicit modeling of two
sexes (both in contact structure and differential transmission
probabilities), temporal and spatial variation in epidemic processes,
or presence of genetic variation in hosts are harder to predict.
